### 2018-06-01 source/Ch2-intro/lofar.tex
added a lot

\pg
LOFAR is a SKA pathfinder instrument, which means that it serves not only as a cutting-edge instrument in its own right, but does so with the explicit aim of serving as testing grounds for technologies \& techniques which could be usefully implemented in the SKA. It is in this context, for example, that trailblazers such as NeNuFar \citepads{2012sf2a.conf..687Z}, a low-frequency extension of LOFAR, are tested. LOFAR is an interferometric array, meaning that it consists of antennas which are combined to form stations, which are themselves distributed throughout the Netherlands and Europe.
\begin{figure}[h!] 
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{images/superterp.png}
	\subcaption{\label{fig.lofar.superterp} LOFAR core, known as the Superterp.}
%    \vspace{4ex}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=.5\linewidth]{images/{lofar-core-map_grey.jpg}}
    \subcaption{\label{fig.lofar.core} Distribution of the so-called LOFAR "core" stations, which include the Superterp.}
%    \vspace{4ex}
  \end{minipage} 
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{images/{Distribution_Remote_Stations}} 
    \subcaption{\label{fig.lofar.remote} LOFAR with both "core" and "remote" stations.}
%    \vspace{4ex}
  \end{minipage}
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{images/{Distribution_International_Stations}} 
    \subcaption{\label{fig.lofar.international} International LOFAR. Newer stations (one in Ireland, three in Poland) are not shown here.}
%    \vspace{4ex}
  \end{minipage} 
\caption{\label{fig.lofar.distribution} Geographic location and distribution of LOFAR stations, explicitly showing what is meant by Superterp, core, remote and international stations. All images from \href{https://www.astron.nl/radio-observatory/astronomers/users/technical-information/lofar-array-configuration/lofar-array-conf}{the official ASTRON website}}
\end{figure}

\pg
There are two bands to LOFAR, which are known as LOFAR-HBA (High-Band Antennas) and LOFAR LBA (Low-Band Antennas). \cref{fig.lofar.superterp,fig.fr606.layout} show the layout of the 6 innermost core stations and the French international LOFAR station, respectively.
\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{images/{LOFAR_NenuFAR.jpg}}
\caption{\label{fig.fr606.layout} Layout of FR606, the French LOFAR station at Nancay. At bottom left are the HBA tiles, bottom right the LBA dipoles, and at the top are some of the NenuFAR mini-arrays.}
\end{figure}

\pg
We see the presence, in both cases, of two very different antenna types. One of these antenna types is not a single antenna, but rather a phased array: 16 antenna dipoles distributed in a $4 \times 4$ array. LOFAR stations include 48 such tiles, distributed in a cross pattern. This pattern is shown in \cref{fig.hba.tile}. Core stations HBA tiles are split into two 24-tile crosses. The antennas from each tile are combined into a phased array with a single ``tile beam", and all tile beams are themselves combined into a station beam\footnote{Referece: \href{https://www.astron.nl/radio-observatory/astronomers/technical-information/antennae/antennae-description}{ASTRON technical description}}. This station beam is then pointed digitally.
\begin{figure}[h!]
\includegraphics[width=0.45\textwidth]{images/{Schematic-diagram-of-a-24-tile-LOFAR-HBA-station-A-tile-is-made-of-16-dual-polarization.png}}
\caption{\label{fig.hba.tile} Layout of a single HBA tile. Source: \href{https://www.researchgate.net/profile/Sarod_Yatawatta/publication/281316394/figure/fig1/AS:614002589720576@1523401033149/Schematic-diagram-of-a-24-tile-LOFAR-HBA-station-A-tile-is-made-of-16-dual-polarization.png}{Sarod Yatawatta researchgate profile}}
\end{figure}

\pg
LOFAR-HBA is sensitive to higher frequencies, from 120 MHz to 240 MHz\footnote{Reference: \href{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}{LOFAR.org website}}. It has a smaller field of view than LOFAR-LBA, but a better resolution.

\pg
LBA antennas, meanwhile, follow a very simple - and cost-effective - design. A typical LBA antenna is shown in \cref{fig.lba.ant}. 96 such antennas are spread in a semi-random pattern in each LOFAR station\footnote{Reference: \href{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}{LOFAR.org website.}}. In survey observations, they are combined as a single phased array.
\begin{figure}[h!]
\includegraphics[width=0.45\textwidth]{images/{lba.png}}
\caption{\label{fig.lba.ant} Picture of a single LBA antenna. Source: \href{https://i2.wp.com/lofar.ie/wp-content/uploads/2017/04/lba.png?resize=1500\%2C1000}{LOFAR technology website.}}
\end{figure}

\pg
The dipole design frees observers from the need to physically point antennas at all: the final station pointing is achieved by digitally introducing delays in observed phase before averaging the data. In this sense, the pointing is achieved in the same way as individual HBA tile pointing, but with one less degree of complexity. These antennas are receptive to signals emitted in the 30-80 MHz frequency range.

>>>>

\pg
LOFAR is a SKA pathfinder instrument, which means that it serves not only as a cutting-edge instrument in its own right, but does so with the explicit aim of serving as testing grounds for technologies \& techniques which could be usefully implemented in the SKA. It is in this context, for example, that trailblazers such as NenuFAR \citepads{2012sf2a.conf..687Z}, a low-frequency extension of LOFAR, are tested. LOFAR is an interferometric array, with each of its elements made up of antennas combined into a phase array. It thus consists of antennas which are combined to form stations, which are themselves distributed throughout the Netherlands and Europe.
\begin{figure}[h!] 
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{images/superterp.png}
	\subcaption{\label{fig.lofar.superterp} LOFAR core, known as the Superterp.}
%    \vspace{4ex}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=.5\linewidth]{images/{lofar-core-map_grey.jpg}}
    \subcaption{\label{fig.lofar.core} Distribution of the so-called LOFAR "core" stations, which include the Superterp.}
%    \vspace{4ex}
  \end{minipage} 
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{images/{Distribution_Remote_Stations}} 
    \subcaption{\label{fig.lofar.remote} LOFAR with both "core" and "remote" stations.}
%    \vspace{4ex}
  \end{minipage}
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{images/{Distribution_International_Stations}} 
    \subcaption{\label{fig.lofar.international} International LOFAR. Newer stations (one in Ireland, three in Poland) are not shown here.}
%    \vspace{4ex}
  \end{minipage} 
\caption{\label{fig.lofar.distribution} Geographic location and distribution of LOFAR stations, explicitly showing what is meant by Superterp, core, remote and international stations. All images from \href{https://www.astron.nl/radio-observatory/astronomers/users/technical-information/lofar-array-configuration/lofar-array-conf}{the official ASTRON website}}
\end{figure}

\pg
There are two bands to LOFAR, which are known as LOFAR-HBA (High-Band Antennas) and LOFAR LBA (Low-Band Antennas). \cref{fig.lofar.superterp,fig.fr606.layout} show the layout of the 6 innermost core stations and the French international LOFAR station, respectively.
\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{images/{LOFAR_NenuFAR.jpg}}
\caption{\label{fig.fr606.layout} Layout of FR606, the French LOFAR station at Nancay. At bottom left are the HBA tiles, bottom right the LBA dipoles, and at the top are some of the NenuFAR tiles.}
\end{figure}

\pg
We see the presence, in both cases, of two very different antenna types. One of these antenna types is not a single antenna, but rather a phased array: 16 antenna dipoles distributed in a $4 \times 4$ array. International LOFAR stations include 48 such tiles, distributed in a ``filled-disk" pattern, where the disk fill ratio is determined by the size of the tile. This pattern is shown in \cref{fig.hba.tile}. Dutch ``remote" stations have 96 such tiles spread in 2 disks (HBA\textunderscore INNER and HBA\textunderscore OUTER), while ``core" station HBA tiles are split into two 24-tile crosses. The antennas from each tile are combined into a phased array with a single ``tile beam", and all tile beams are themselves combined as a phased array into a station beam\footnote{Referece: \href{https://www.astron.nl/radio-observatory/astronomers/technical-information/antennae/antennae-description}{ASTRON technical description}}. This station beam is then pointed digitally.
\begin{figure}[h!]
\includegraphics[width=0.45\textwidth]{images/{Schematic-diagram-of-a-24-tile-LOFAR-HBA-station-A-tile-is-made-of-16-dual-polarization.png}}
\caption{\label{fig.hba.tile} Layout of a single HBA tile. Source: \href{https://www.researchgate.net/profile/Sarod_Yatawatta/publication/281316394/figure/fig1/AS:614002589720576@1523401033149/Schematic-diagram-of-a-24-tile-LOFAR-HBA-station-A-tile-is-made-of-16-dual-polarization.png}{Sarod Yatawatta researchgate profile}}
\end{figure}

\pg
LOFAR-HBA is sensitive to higher frequencies, from 120 MHz to 240 MHz\footnote{Reference: \href{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}{LOFAR.org website}}. It has a smaller field of view than LOFAR-LBA, but a better angular resolution.

\pg
LBA antennas, meanwhile, follow a very simple - and cost-effective - design. A typical LBA antenna is shown in \cref{fig.lba.ant}. 96 such antennas are spread in a semi-random pattern in each LOFAR station\footnote{Reference: \href{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}{LOFAR.org website.}}. In Dutch stations, 48 of these are used at all times. They are combined as a single phased array.
\begin{figure}[h!]
\includegraphics[width=0.45\textwidth]{images/{lba.png}}
\caption{\label{fig.lba.ant} Picture of a single LBA antenna. Source: \href{https://i2.wp.com/lofar.ie/wp-content/uploads/2017/04/lba.png?resize=1500\%2C1000}{LOFAR technology website.}}
\end{figure}

\pg
The dipole design frees observers from the need to physically point antennas at all: the final station pointing is achieved by digitally introducing delays in observed phase before averaging the data. In this sense, the pointing is achieved in a similar way as individual HBA tile pointing, albeit in analog. These antennas are receptive to signals emitted in the 30-80 MHz frequency range.

### 2018-06-01 source/Ch2-intro/science.tex
added a lot

\pg

@@

\subsection{Free-Free Radiation}
\pg
Free-free or ``bremsstrahlung" (``braking") radiation occurs when the trajectory of a high-energy charged particle is deflected by an electric field. This non-thermal emission mechanism is the dominant mechanism in HII regions (which contain ionised hydrogen), where star formation has previously taken place. It is called free-free emission because it is produced by free electrons scattering off ionised hydrogen without being captured. % as shown in Fig. \ref{plot.freefree}
%\begin{figure*}[!h]
%\centering
%\includegraphics[width=0.3\textwidth]{images/bremsstrahlung.png}
%\caption{\label{plot.freefree} Schematic showing the physical process leading to free-free emission. As we can see, the electron is scattered, but not captured. Source: \url{https://thephysicsbehind.com/2015/04/16/x-ray-tubes/}}
%\end{figure*}

\pg
This emission's spectrum is heavily dependent on a number of factors: including frequency, temperature, and critically, free-free opacity $\tau_\nu$, itself a function of electron density. It is characterised by a knee in its spectrum, occurring where $\tau_\nu\sim 1$. This knee delineates two regions with different spectral indices; $\alpha \sim -0.1$ at higher frequencies, and $\alpha \leq 2$ at lower frequencies. This gives a characteristic shape, shown in Fig. \ref{plot.freefree.spectrum}.
\begin{figure*}[!h]
\centering
\includegraphics[width=0.9\textwidth]{images/freefree.png}
\caption{\label{plot.freefree.spectrum} Characteristic spectrum of free-free radiation. Arbitrary unit scale. Source: NRAO online course.}
\end{figure*}

\pg
As we can see, this spectrum falls off sharply with frequency. As such, while it is expected to dominate over thermal emission in the absence of synchrotron radiation, it is unlikely to dominate in the LOFAR band. It acts as a tracer for star formation and other "gentler" physical processes detected at low radio frequencies.

\subsection{Synchrotron Radiation}
\pg
Synchrotron radiation (or "magnetobremsstrahlung") occurs when the trajectory of a high-energy charged particle is deflected by a magnetic field. As the German name suggests, it is the magnetic equivalent of free-free radiation. It is the tracer of extremely violent processes, such as AGN jets. In its mildly relativistic regime, it is referred to as cyclotron radiation, after the device in which it was first tested, and in non-relativistic regimes, it is known as gyro radiation.

\pg

>>>>

\subsection{Synchrotron Radiation}

\pg
Synchrotron radiation is the dominant mode of emission for extragalactic sources at low wavelengths. As such, it will be covered in greater detail that the other mechanisms above, which are important in other bands (and therefore relevant for multi-spectral analysis) but much less relevant to LOFAR observations. This section draws heavily from Garret Cotter's \href{http://www-astro.physics.ox.ac.uk/~garret/teaching/}{high-energy astrophysics lectures}, as well as Alan Loh's \href{http://theses.md.univ-paris-diderot.fr/LOH_Alan_2_va_20160930.pdf}{doctoral thesis} and \href{https://www.cv.nrao.edu/course/astr534/SelfAbsorption.html}{the NRAO online course}.

\pg
Synchrotron radiation (or "magnetobremsstrahlung") occurs when the trajectory of a high-energy charged particle is deflected by a magnetic field. As the German name suggests, it is the magnetic equivalent of free-free radiation. It is the tracer of extremely violent processes, such as AGN jets. In the low-energy regime, it is also known as cyclotron radiation, after the device in which it was first measured in a laboratory. %, and in non-relativistic regimes, it is known as gyro radiation.

\pg
The synchrotron power spectrum of a population of electrons with exactly the same energy is a function of its Lorentz factor and its emission angle\footnote{The emission angle is the angle between the electron's velocity and its radial velocity.} \citepads{1994hea2.book.....L}. The typical radiation spectrum for a single electron is shown in \cref{fig.synchrotron.1electrion} as a function of the critical frequency, which is itself a function of the emission angle and the synchrotron frequency\footnote{Which is itself a function of the Lorentz factor and the magnetic field strength - $\nu_\mathrm{sync}=\frac{\gamma^2 e B}{2\pi m_e c}$ .}: $\nu_c=3/2\nu_\mathrm{sync}\sin(\theta)$.
\begin{figure}[!h]
\centering
\includegraphics[width=0.9\textwidth]{images/Synchrotron-single-electron.png}
\caption{\label{fig.synchrotron.1electrion} Characteristic spectrum of synchrotron emission for a population of single-energy electrons. Arbitrary unit scale. Source: NRAO online course.}
\end{figure}

\pg
Of course, in reality, we know that electron populations in astrophysical sources do not all share exactly the same energy when emitting. The effect of the electron energy distribution must therefore be taken into account. Because the population in question is energised through non-thermal processes, its energy distribution is not Maxwellian, and is generally written as a power law: $S_\nu \propto \nu^{\alpha}$, where $\alpha$ is the object's spectral index. %If the dominant mechanism for charged particle acceleration is Fermi acceleration (see \citetads{1978MNRAS.182..147B} and companion papers, \citetads{2007Ap&SS.309..119R}), aggregate spectrum can be described as a power law in Lorentz factor:
%\begin{align}
%P_\mathrm{sync} (\nu,\theta) &= \kappa B^{.5(p+1)} \nu^{-.5(p-1)}
%\end{align} 
The question now becomes: what $\alpha$ is appropriate for what kind of object? Does every source have the same $\alpha$ at all frequencies? Do all synchrotron sources have the same spectral index at a given frequency, and if not, what does this tell us?

\pg
To answer these questions, we must recall that so far, we have assumed that every photon emitted through synchrotron radiation reaches us. This is not necessarily the case in practice: it is possible for a given photon to be scattered (i.e. absorbed \& re-emitted) by surrounding synchrotron electrons. The likelihood of this occurring at a given frequency is a function of the absorption cross-section, and is a complex quantity to calculate (see \citetads{1994hea2.book.....L} pp258-260). Suffice to say that at longer wavelengths (lower frequencies) absorption is more effective, and at shorter wavelengths (higher frequencies) the underlying electron energy power-law  distribution can be seen. An example is shown in \cref{fig.synchrotron}, taken from \citetads{2007A&A...471.1105T,}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{images/synchrotron-spectrum.png}
\caption{\label{fig.synchrotron} Power spectrum for J0229.1-0402.}
\end{figure}

\pg
In the optically-thick regime, we know that the most effective radiation mechanism is simple black-body radiation, and so the power spectrum at a given frequency can be defined in terms of an ``electron temperature". In the Rayleigh-Jeans approxmation (valid as we are explicitly in the low-frequency regime), this gives $S_\nu \propto \nu{5/2}$, or $\alpha=5/2$. 
In the optically-thin regime, meanwhile, we recover a mixture of the underlying power-law distribution ($S_\nu \propto e^{-\nu}$) and black-body radiation (which has a frequency-dependent $\alpha$). Empirically, we generally observe $\alpha \sim 0.7$.

@@

\subsection{Free-Free Radiation}
\pg
Free-free or ``bremsstrahlung" (``braking") radiation occurs when the trajectory of a charged particle is deflected by an electric field. This non-thermal emission mechanism is the dominant mechanism in HII regions (which contain ionised hydrogen), where star formation has previously taken place. It is called free-free emission because it is produced by free electrons deviating off ionised hydrogen without being captured. % as shown in Fig. \ref{plot.freefree}
%\begin{figure*}[!h]
%\centering
%\includegraphics[width=0.3\textwidth]{images/bremsstrahlung.png}
%\caption{\label{plot.freefree} Schematic showing the physical process leading to free-free emission. As we can see, the electron is scattered, but not captured. Source: \url{https://thephysicsbehind.com/2015/04/16/x-ray-tubes/}}
%\end{figure*}

\pg
This emission's spectrum is heavily dependent on a number of factors: including frequency, temperature, and critically, free-free opacity $\tau_\nu$, itself a function of electron density. It is characterised by a knee in its spectrum, occurring where $\tau_\nu\sim 1$. This knee delineates two regions with different spectral indices; $\alpha \sim -0.1$ at higher frequencies, and $\alpha \leq 2$ at lower frequencies. This gives a characteristic shape, shown in Fig. \ref{plot.freefree.spectrum}.
\begin{figure*}[!h]
\centering
\includegraphics[width=0.8\textwidth]{images/freefree.png}
\caption{\label{plot.freefree.spectrum} Characteristic spectrum of free-free radiation. Arbitrary unit scale. Source: NRAO online course.}
\end{figure*}

\pg
As we can see, this spectrum falls off sharply with frequency. As such, while it is expected to dominate over thermal emission in the absence of synchrotron radiation, it is synchrotron emission that is expected to dominate overall - if present - in the LOFAR band. Free-free emission acts as a tracer for star formation and other "gentler" physical processes detected at low radio frequencies.

### 2018-06-01 source/Ch3-RIME/RIME.tex
added a lot

\chapter{Mathematical Framework: the RIME Formalism}
\pg
In this section, we introduce the mathematical framework which forms the basis for our algorithmic work.
\label{section.RIME}
While radio interferometry has historically been a complex business \citepads{2001isra.book.....T}, modern radio interferometry is based on the Radio Interferometer's Measurement Equation, a powerful and elegant formalism which underpins modern calibration and imaging algorithms. In this section, I will give a simplified account of this formalism as described in \citetads{2011A&A...527A.106S} and its companion papers\footnote{\citetads{2011A&A...527A.107S}, \citetads{2011A&A...527A.108S} and \citetads{2011A&A...531A.159S}}. This will form the theoretical basis on which further sections will expand. I cannot overstate the importance of the RIME as the foundation of modern calibration and imaging algorithms.

@@

\pg
Consider a single quasi-monochromatic point source in the sky. Its signal at a point in time and space can be described by a complex vector $\evector$. We can then represent physical processes which affect this signal's propagation using \emph{Jones matrices}. In other words:
\begin{equation}
\volt = \Jones\evector
\end{equation}
where $\volt$ is the voltage measured by our antenna, and $\Jones$ the Jones matrix describing the \emph{net} propagation effects which affect our source's signal. $\Jones$ can be written as a series of matrix products, each individual matrix describing a physical phenomenon. They are then called a \emph{Jones chain}, and the resulting $\Jones$ is often called the \emph{total Jones matrix}.

>>>>

\chapter{Mathematical Framework: the RIME Formalism}\label{section.RIME}
\minitoc

\pg
In this section, we introduce the mathematical framework which forms the basis for our algorithmic work.
While radio interferometry has historically been a complex business \citepads{2001isra.book.....T}, modern radio interferometry is based on the Radio Interferometer's Measurement Equation, a powerful and elegant formalism which underpins modern calibration and imaging algorithms. In this section, I will give a simplified account of this formalism as described in \citetads{2011A&A...527A.106S} and its companion papers\footnote{\citetads{2011A&A...527A.107S}, \citetads{2011A&A...527A.108S} and \citetads{2011A&A...531A.159S}}. This will form the theoretical basis on which further sections will expand. I cannot overstate the importance of the RIME as the foundation of modern calibration and imaging algorithms.

@@

\pg
Consider a single quasi-monochromatic point source in the sky. Its signal at a point in time and space can be described by a complex vector $\evector$. We can then represent physical processes which affect this signal's propagation using \emph{Jones matrices}. In other words:
\begin{equation}
\volt = \Jones\evector
\end{equation}
where $\volt$ is the voltage measured by our antenna, and $\Jones$ the Jones matrix describing the \emph{net} propagation effects - atmospheric, astrophysical, and instrumental - which affect our source's signal. $\Jones$ can be written as a series of matrix products, each individual matrix describing a physical phenomenon. They are then called a \emph{Jones chain}, and the resulting $\Jones$ is often called the \emph{total Jones matrix}.
